---
author: "CC"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    rmdformats::material:
        use_bookdown: true
        thumbnails: false
        df_print: kable
        code_folding: hide
        number_sections: yes
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
title: |
  | Step 3.1: Cell populations orverview -- `r DATASET`
---
  
  <!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->
  
  <!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
  <!-- <script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->
<!--
  <style>
  .zoomDiv {
    display: none;
    position: fixed;
    top: 50%;
    left: 50%;
    z-index: 50;
    transform: translate(-50%, -50%);
    background-color: #FFFFFF;
      box-shadow: 0px 0px 50px #888888;
    width: fit-content;
    max-width: 90%;
    max-height: 90%;
    overflow: auto;
  }

.zoomImg {
  width: 150%;
}
</style>
    <script type = "text/javascript">
  $(document).ready(function() {
    $("body").prepend("<div class = \"zoomDiv\"><img src = \"\" class = \"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use "img.zoom" and out.extra = "class = \"zoom\"" in chunk to specify manually which chunk images can be zoomed
    $("img:not(.zoomImg)").filter(":not(a *)").click(function() {
      $(".zoomImg").attr("src", $(this).attr("src"));
      $(".zoomDiv").show();
    })
    // onClick function for hiding div
    $("img.zoomImg").click(function() {
      $(".zoomDiv").hide();
    })
  })
</script> -->

  
```{r setup, include = FALSE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,
  
  # text output
  echo = TRUE,
  results = "hold",
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,
  
  # code decoration
  tidy.opts = list(width.cutoff = 90),
  comment = "",
  attr.output = ".numberLines",
  
  # plots
  fig.path = paste0(PATH_OUT_FIG, "/03_1_manualControls_"),      # is set later, in chunk setup2
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c("png", "pdf"),
  fig.width = 12,
  fig.height = 12,
  fig.align = "center"   # should be tuned to default in multiple plots chunk
)
```

```{r load-libraries, include = FALSE}
library(kableExtra)
library(dplyr)
library(Seurat)
library(stringr)
library(ggplot2)

source("./plotting_functions.R")
```

```{r dir-managment, include = FALSE}
PATH_OUT_ANALYSIS <- file.path(PATH_ROOT, "03_1_manualControls")
if (!dir.exists(PATH_OUT_ANALYSIS)) { dir.create(PATH_OUT_ANALYSIS) }
```

# Load `r DATASET` data

```{r load-data}
cplt <- readRDS(paste0(PATH_RDS_OBJECTS, "/02_preprocessed_", DATASET, "_complete.rds"))
metadata <- read.table(paste0(PATH_ROOT, "/03_qc/metadata_", DATASET, "_complete.csv"), header = TRUE, sep = ",")
umapCoord <- read.table(paste0(PATH_ROOT, "/04_process/DR_umap_", DATASET, "_complete.csv"), header = TRUE, sep = ",")

# reintegrate umap coordinates
umapMat <- as.matrix(umapCoord)
cplt[["umap"]] <- CreateDimReducObject(embeddings = umapMat, key = "UMAP_")

# reintegrate metadata
cplt <- AddMetaData(object = cplt, metadata = metadata$percent.mito,   col.name = "percent.mito")
cplt <- AddMetaData(object = cplt, metadata = metadata$percent.ribo,   col.name = "percent.ribo")
```

# Mitochondrial exploration 

```{r define-groups-mito, out.width="50%", fig.align='default', fig.show='hold'}
cplt@meta.data[,"mito_groups"] <- set_thresh(mito_thresholds, cplt@meta.data$percent.mito, "mito")
# Must add an upper bound in order for the last group to be displayed
max_mito <- max(cplt@meta.data$percent.mito)
mito_thresholds <- c(0,mito_thresholds)

plot1 <- VlnPlot(cplt, features = c("percent.mito")) +
  lapply(mito_thresholds, threshline) +
  ggtitle("Mitochondrial expression\npercentage") +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(angle = 0, size = 24, hjust = 0.5, face = "bold"),
        text = element_text(size = 24),
        plot.subtitle = element_text(hjust = 0.5)) +
  NoLegend() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  coord_cartesian(ylim = c(0, 100)) +
  mapply(threshtext, mito_thresholds, seq_along(mito_thresholds), "mito")
plot2 <- plot1 +
    ggtitle("Zoom in") +
    coord_cartesian(ylim = c(0, 15))

plot1
plot2
```

Groups' size:
```{r groups-size-mito}
mito_groups <- sort(unique(cplt$mito_groups))
check_empty_groups(cplt$mito_groups, mito_groups)
table(cplt@meta.data$mito_groups)
```

```{r groups-mito, out.width="50%", fig.align='default', fig.show='hold'}
cellsData <- data.frame(umapCoord, metadata$percent.mito, cplt@meta.data$mito_groups)
colnames(cellsData) <- c(colnames(umapCoord), "percent.mito", "group_mito")

ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( data = cellsData, # Now provide data including column for facetting
                aes(color = percent.mito),
                alpha = 0.6,
                size  = 1.2) +
    NoLegend() +
    ggtitle(paste0(DATASET, "\nmitochondrial gene expression level")) +
    defaultTheme() +
    theme(legend.text = element_text(size = 10),
          legend.title = element_text(size = 14))

mito_labels <- threshlabel(mito_thresholds)

ggplot(data = cellsData,
    aes(x = UMAP_1,
        y = UMAP_2,
        colour = factor(group_mito))) +
    geom_point() +
    gghighlight::gghighlight() +
    facet_wrap(vars(group_mito)) +
    ggtitle(paste0(DATASET, "\nmitochondrial gene expression groups")) +
    defaultTheme() +
    scale_color_discrete(name = "percent.mito (%)",
                      labels = mito_labels) +
    theme(strip.text.x = element_text(size = 20),
          legend.text = element_text(size = 20),
          legend.title = element_text(size = 28),
        legend.position = "bottom",
        legend.direction = "vertical")
```

# Ribosomal exploration 

```{r define-groups-ribo, out.width="50%"}
cplt@meta.data[, "ribo_groups"] <- set_thresh(ribo_thresholds, cplt@meta.data$percent.ribo, "ribo")
max_ribo <- max(cplt@meta.data$percent.ribo)
ribo_thresholds <- c(0,ribo_thresholds)

VlnPlot(cplt, features = c("percent.ribo")) +
    lapply(ribo_thresholds, threshline) +
    ggtitle("Ribosomal expression\npercentage") +
    theme(axis.title.x = element_blank(),
          axis.text = element_text(size = 16),
          axis.text.x = element_text(angle = 0, size = 24, hjust = 0.5, face = "bold"),
          text = element_text(size = 24),
          plot.subtitle = element_text(hjust = 0.5)) +
    NoLegend() +
    scale_y_continuous(labels = scales::percent_format(scale = 1)) +
    mapply(threshtext, ribo_thresholds, seq_along(ribo_thresholds), "ribo")
```

Groups' size:
```{r groups-size-ribo}
ribo_groups <- sort(unique(cplt$ribo_groups))
check_empty_groups(cplt$ribo_groups, ribo_groups)
table(cplt@meta.data$ribo_groups)
```

```{r groups-ribo, out.width="50%", fig.align='default', fig.show='hold'}
cellsData <- data.frame(umapCoord, metadata$percent.ribo, cplt@meta.data$ribo_groups)
colnames(cellsData) <- c(colnames(umapCoord), "percent.ribo", "group_ribo")

ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP_1,
            y = UMAP_2)) +
  geom_point( data = cellsData, # Now provide data including column for facetting
              aes(color = percent.ribo),
              alpha = 0.6,
              size  = 1.2) +
  NoLegend() +
  ggtitle(paste0(DATASET, "\nribosomal gene expression level")) +
  defaultTheme() +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 14))

ribo_labels <- threshlabel(ribo_thresholds)

ggplot(data = cellsData,
       aes(x = UMAP_1,
           y = UMAP_2,
           colour = factor(group_ribo))) +
  geom_point() +
  gghighlight::gghighlight() +
  facet_wrap(vars(group_ribo),
             ncol = 2) +
  ggtitle(paste0(DATASET, "\nribosomal gene expression groups")) +
  defaultTheme() +
  scale_color_discrete(name = "percent.ribo (%)",
                       labels = ribo_labels) +
  theme(strip.text.x = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 28),
        legend.position = "bottom",
        legend.direction = "vertical")
```


# UMI read counts exploration 

```{r define-groups-UMI, out.width="50%"}
cplt@meta.data[, "umi_groups"] <- set_thresh(umi_thresholds, cplt@meta.data$nCount_RNA, "umi")

umi_thresholds <- c(0,umi_thresholds)

VlnPlot(cplt, features = c("nCount_RNA")) +
  lapply(umi_thresholds, threshline) +
  ggtitle("UMI read counts") +
  theme(axis.title.x = element_blank(),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(angle = 0, size = 24, hjust = 0.5, face = "bold"),
        text = element_text(size = 24),
        plot.subtitle = element_text(hjust = 0.5)) +
  NoLegend() +
  mapply(threshtext, umi_thresholds, seq_along(umi_thresholds), "umi")
```

Groups' size:
```{r groups-size-UMI}
umi_groups <- sort(unique(cplt$umi_groups))
check_empty_groups(cplt$umi_groups, umi_groups)
table(cplt@meta.data$umi_groups)
```

```{r groups-UMI, out.width="50%", fig.align='default', fig.show='hold'}
cellsData <- data.frame(umapCoord, cplt@meta.data$nCount_RNA, cplt@meta.data$umi_groups)
colnames(cellsData) <- c(colnames(umapCoord), "nCount_RNA", "group_umi")

ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( data = cellsData, # Now provide data including column for facetting
                aes(color = nCount_RNA),
                alpha = 0.6,
                size  = 1.2) +
    NoLegend() +
    ggtitle(paste0(DATASET, "\nUMI read counts")) +
    defaultTheme() +
    theme(legend.text = element_text(size = 10),
          legend.title = element_text(size = 14))

umi_labels <- threshlabel(umi_thresholds)

ggplot(data = cellsData,
    aes(x = UMAP_1,
        y = UMAP_2,
        colour = factor(group_umi))) +
    geom_point() +
    gghighlight::gghighlight() +
    facet_wrap(vars(group_umi),
               ncol = 2) +
    ggtitle(paste0(DATASET, "\nUMI read counts groups")) +
    defaultTheme() +
    scale_color_discrete(name = "UMI read counts",
                      labels = umi_labels) +
    theme(strip.text.x = element_text(size = 20),
          legend.text = element_text(size = 20),
          legend.title = element_text(size = 28),
        legend.position = "bottom",
        legend.direction = "vertical")
```



# Feature counts exploration 

```{r define-groups-feature, out.width="50%"}
cplt@meta.data[, "feature_groups"] <- set_thresh(feature_thresholds, cplt@meta.data$nFeature_RNA, "feature")
max_feature <- max(cplt@meta.data$nFeature_RNA)
feature_thresholds <- c(0,feature_thresholds)

VlnPlot(cplt, features = c("nFeature_RNA")) +
    lapply(feature_thresholds, threshline) +
    ggtitle("feature read counts") +
    theme(axis.title.x = element_blank(),
          axis.text = element_text(size = 16),
          axis.text.x = element_text(angle = 0, size = 24, hjust = 0.5, face = "bold"),
          text = element_text(size = 24),
          plot.subtitle = element_text(hjust = 0.5)) +
    NoLegend() +
    mapply(threshtext, feature_thresholds, seq_along(feature_thresholds), "feature")
```

Groups' size:
```{r groups-size-feature}
feature_groups <- sort(unique(cplt$feature_groups))
check_empty_groups(cplt$feature_groups, feature_groups)
table(cplt@meta.data$feature_groups)
```

```{r groups-feature, out.width="50%", fig.align="default", fig.show="hold"}
cellsData <- data.frame(umapCoord, cplt@meta.data$nFeature_RNA, cplt@meta.data$feature_groups)
colnames(cellsData) <- c(colnames(umapCoord), "nFeature_RNA", "group_feature")

ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
       aes( x = UMAP_1,
            y = UMAP_2)) +
  geom_point( data = cellsData, # Now provide data including column for facetting
              aes(color = nFeature_RNA),
              alpha = 0.6,
              size  = 1.2) +
  NoLegend() +
  ggtitle(paste0(DATASET, "\nfeature counts")) +
  defaultTheme() +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 14))

feature_labels <- threshlabel(feature_thresholds)

ggplot(data = cellsData,
       aes(x = UMAP_1,
           y = UMAP_2,
           colour = factor(group_feature))) +
  geom_point() +
  gghighlight::gghighlight() +
  facet_wrap(vars(group_feature),
             ncol = 2) +
  ggtitle(paste0(DATASET, "\nfeature read counts groups")) +
  defaultTheme() +
  scale_color_discrete(name = "feature read counts",
                       labels = feature_labels) +
  theme(strip.text.x = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 28),
        legend.position = "bottom",
        legend.direction = "vertical")
```

# Cross-validations

```{r cross-validation}
allGroups <- c("mito_groups", "ribo_groups", "umi_groups", "feature_groups")
invisible(apply(combn(allGroups, 2), 2, function(x) {
  cat(paste0(x[1], " and ", x[2], " co-ocurrence"))
  print(table(cplt@meta.data[,x[1]], cplt@meta.data[,x[2]]))
  cat("\n\n")
}))

mito_text <- ""
for(i in 1:length(mito_labels)){
  mito_text <- paste0(mito_text, " - gp", i, "_mito&emsp;", mito_labels[i],"\n")
}
```

## Mitochondrial levels CV {.tabset .tabset-pills .tabset-fade}

The various levels of mitochondrial expression represent (values in %):

`r mito_text`


```{r cross-val-mito, results='asis', out.width='50%', fig.align='default'}
vars <- allGroups[allGroups != "mito_groups"]

invisible(sapply(mito_groups, function(group) {
  
  cat("### ", group, "\n")
  
  Idents(cplt) <- "mito_groups"
  
  # Highlight cells of current cluster on a dimreduc plot
  highlightClusterPlot(group, seuratObject = cplt, reduction = ifelse( exists("useReduction"), useReduction, "umap"))
  
  invisible(sapply(vars, function(vrbl) {
    cellsData <- data.frame(cplt@reductions$umap@cell.embeddings, cplt@meta.data$mito_groups, cplt@meta.data[[vrbl]])
    colnames(cellsData) <- c(colnames(umapCoord), "mito_groups", vrbl)
    
    print(
      ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
             aes( x = UMAP_1,
                  y = UMAP_2)) +
        geom_point( alpha = .4,
                    size  = 1,
                    color = "grey") +
        geom_point( data = cellsData[cellsData$mito_groups == group,],
                    aes(color = get(vrbl)),
                    alpha = .5,
                    size  = 1.2) +
        scale_color_discrete(name = vrbl) +
        NoLegend() +
        ggtitle(paste0(DATASET, "\n", group, " versus ", vrbl)) +
        defaultTheme() +
        theme(legend.text = element_text(size = 15),
              legend.title = element_text(size = 20)) +
        guides(colour = guide_legend(override.aes = list(size=10)))
    )
  }))
  
  cat(" \n \n") # Required for '.tabset'
}))

ribo_text <- ""
for(i in 1:length(ribo_labels)){
  ribo_text <- paste0(ribo_text, " - gp", i, "_ribo&emsp;", ribo_labels[i],"\n")
}
```

## Ribosomal levels CV {.tabset .tabset-pills .tabset-fade}

The various levels of ribosomal expression represent (values in %):
  
`r ribo_text`

```{r cross-val-ribo, results='asis', out.width='50%', fig.align='default'}
vars <- allGroups[allGroups != "ribo_groups"]

invisible(sapply(ribo_groups, function(group) {
  
  cat("### ", group, "\n")
  
  Idents(cplt) <- "ribo_groups"
  
  # Highlight cells of current cluster on a dimreduc plot
  highlightClusterPlot(group, seuratObject = cplt, reduction = ifelse( exists("useReduction"), useReduction, "umap"))
  
  invisible(sapply(vars, function(vrbl) {
    cellsData <- data.frame(cplt@reductions$umap@cell.embeddings, cplt@meta.data$ribo_groups, cplt@meta.data[[vrbl]])
    colnames(cellsData) <- c(colnames(umapCoord), "ribo_groups", vrbl)
    
    print(
      ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
             aes( x = UMAP_1,
                  y = UMAP_2)) +
        geom_point( alpha = .4,
                    size  = 1,
                    color = "grey") +
        geom_point( data = cellsData[cellsData$ribo_groups == group,],
                    aes(color = get(vrbl)),
                    alpha = .5,
                    size  = 1.2) +
        scale_color_discrete(name = vrbl) +
        NoLegend() +
        ggtitle(paste0(DATASET, "\n", group, " versus ", vrbl)) +
        defaultTheme() +
        theme(legend.text = element_text(size = 15),
              legend.title = element_text(size = 20)) +
        guides(colour = guide_legend(override.aes = list(size=10)))
    )
  }))
  
  cat(" \n \n") # Required for '.tabset'
}))

umi_text <- ""
for(i in 1:length(umi_labels)){
  umi_text <- paste0(umi_text, " - gp", i, "_umi&emsp;", umi_labels[i],"\n")
}
```


## UMI read counts levels CV {.tabset .tabset-pills .tabset-fade}

The various levels of UMI read counts are:

`r umi_text`

```{r cross-val-umi, results='asis', out.width='50%', fig.align='default'}
vars <- allGroups[allGroups != "umi_groups"]

invisible(sapply(umi_groups, function(group) {
  
  cat("### ", group, "\n")
  
  Idents(cplt) <- "umi_groups"
  
  # Highlight cells of current cluster on a dimreduc plot
  highlightClusterPlot(group, seuratObject = cplt, reduction = ifelse( exists("useReduction"), useReduction, "umap"))
  
  invisible(sapply(vars, function(vrbl) {
    cellsData <- data.frame(cplt@reductions$umap@cell.embeddings, cplt@meta.data$umi_groups, cplt@meta.data[[vrbl]])
    colnames(cellsData) <- c(colnames(umapCoord), "umi_groups", vrbl)
    
    print(
      ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
             aes( x = UMAP_1,
                  y = UMAP_2)) +
        geom_point( alpha = .4,
                    size  = 1,
                    color = "grey") +
        geom_point( data = cellsData[cellsData$umi_groups == group,],
                    aes(color = get(vrbl)),
                    alpha = .5,
                    size  = 1.2) +
        scale_color_discrete(name = vrbl) +
        NoLegend() +
        ggtitle(paste0(DATASET, "\n", group, " versus ", vrbl)) +
        defaultTheme() +
        theme(legend.text = element_text(size = 15),
              legend.title = element_text(size = 20)) +
        guides(colour = guide_legend(override.aes = list(size=10)))
    )
  }))
  
  cat(" \n \n") # Required for '.tabset'
}))

feature_text <- ""
for(i in 1:length(feature_labels)){
  feature_text <- paste0(feature_text, " - gp", i, "_feature&emsp;", feature_labels[i],"\n")
}
```


## Feature counts levels CV {.tabset .tabset-pills .tabset-fade}

The various levels of feature counts are:
  
`r feature_text`

```{r cross-val-feature, results='asis', out.width='50%', fig.align='default'}
vars <- allGroups[allGroups != "feature_groups"]

invisible(sapply(feature_groups, function(group) {
  
  cat("### ", group, "\n")
  
  Idents(cplt) <- "feature_groups"
  
  # Highlight cells of current cluster on a dimreduc plot
  highlightClusterPlot(group, seuratObject = cplt, reduction = ifelse( exists("useReduction"), useReduction, "umap"))
  
  invisible(sapply(vars, function(vrbl) {
    cellsData <- data.frame(cplt@reductions$umap@cell.embeddings, cplt@meta.data$feature_groups, cplt@meta.data[[vrbl]])
    colnames(cellsData) <- c(colnames(umapCoord), "feature_groups", vrbl)
    
    print(
      ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
             aes( x = UMAP_1,
                  y = UMAP_2)) +
        geom_point( alpha = .4,
                    size  = 1,
                    color = "grey") +
        geom_point( data = cellsData[cellsData$feature_groups == group,],
                    aes(color = get(vrbl)),
                    alpha = .5,
                    size  = 1.2) +
        scale_color_discrete(name = vrbl) +
        NoLegend() +
        ggtitle(paste0(DATASET, "\n", group, " versus ", vrbl)) +
        defaultTheme() +
        theme(legend.text = element_text(size = 15),
              legend.title = element_text(size = 20)) +
        guides(colour = guide_legend(override.aes = list(size=10)))
    )
  }))
  
  cat(" \n \n") # Required for '.tabset'
}))
```


# Exports

```{r exports}
# Groups metadata
keepMDnames <- c("orig.ident", "nCount_RNA", "nFeature_RNA")
cplt@meta.data <- cbind(cellIDs = rownames(cplt@meta.data), cplt@meta.data)
exportMD <- names(cplt@meta.data)[! names(cplt@meta.data) %in% keepMDnames]

write.table(
  cplt@meta.data[exportMD],
  file.path(
    PATH_OUT_ANALYSIS, 
    paste0("metadataGroups_", DATASET, ".csv")),
  row.names = FALSE,
  col.names = TRUE,
  sep = ","
)

invisible(gc())
```

# SessionInfo

```{r sessionInfo}
sessionInfo()
```