---
author: "Céline Chevalier and Anaı̈s Baudot"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
# title: |
#   | Doublets removal
#   | Step {{STEP_ID}}
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type = "text/javascript">
  $(document).ready(function() {
    $("body").prepend("<div class = \"zoomDiv\"><img src = \"\" class = \"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use "img.zoom" and out.extra = "class = \"zoom\"" in chunk to specify manually which chunk images can be zoomed
    $("img:not(.zoomImg)").filter(":not(a *)").click(function() {
      $(".zoomImg").attr("src", $(this).attr("src"));
      $(".zoomDiv").show();
    })
    // onClick function for hiding div
    $("img.zoomImg").click(function() {
      $(".zoomDiv").hide();
    })
  })
</script>


```{r setup, include = FALSE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,

  # text output
  echo = TRUE,
  results = "hold",
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,

  # code decoration
  tidy.opts = list(width.cutoff = 90),
  comment = "",
  attr.output = ".numberLines",

  # plots
  fig.path = paste0(PATH_OUT_FIG, "/06_annotDEAviz_scenario_", scenario, "_"),
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c("png", "pdf"),
  out.width = "50%",
  fig.width = 12,
  fig.height = 12,
  fig.align = "center"   # should be tuned to default in multiple plots chunk
)
```

```{r load-libraries, include = FALSE}
library(Seurat)
library(ggplot2)
library(biomaRt)
library(tidyr)
library(Matrix)
library(dplyr)
library(kableExtra)
library(stringr)
library(DT)
library(scales)
library(RColorBrewer)
library(grDevices)


# source("./extract_top_features.R")
# source("./highlightClusterPlot.R")
# source("./additionalPlots.R")
source("./plotting_functions.R")
source("./data_management.R")
```

```{r load-colors, eval=atlas, include = FALSE}
# Create color vectors
colors.table <- read.csv(file.path(PATH_ATLAS, "colorsSheet.csv"), header=T, comment.char="", as.is=T)
colors.celltype <- setNames(colors.table$celltype_color[!is.na(colors.table$celltype)], colors.table$celltype[!is.na(colors.table$celltype)])
colors.stage <- setNames(colors.table$development_stage_color[!is.na(colors.table$development_stage)], colors.table$development_stage[!is.na(colors.table$development_stage)])
```

```{r dir-managment-1, include=FALSE, eval=combinedD}
# PATHS
PATH_RDS_FILE <- paste0(PATH_RDS_OBJECTS, "/07_preprocessed_", DATASET, "_scenario.", scenario, ".rds")
PATH_CLUSTER_TABLE <- paste0(PATH_ROOT, "/08_combineData/clustering_", DATASET, "_scenario.", scenario, "_meth.", clust_meth, "_res.", clust_res, ".csv")
PATH_OUT_ANALYSIS <- file.path(PATH_ROOT, "08_combineData")
if (!dir.exists(PATH_OUT_ANALYSIS)) { dir.create(PATH_OUT_ANALYSIS) }
```

```{r dir-managment-2, include=FALSE, eval=!combinedD}
# PATHS
PATH_RDS_FILE <- paste0(PATH_RDS_OBJECTS, "/05_preprocessed_", DATASET, "_scenario.", scenario, ".rds")
PATH_CLUSTER_TABLE <- paste0(PATH_ROOT, "/06_process/clustering_", DATASET, "_scenario.", scenario, "_meth.", clust_meth, "_res.", clust_res, ".csv")
PATH_OUT_ANALYSIS <- file.path(PATH_ROOT, "07_annotDEAviz")
if (!dir.exists(PATH_OUT_ANALYSIS)) { dir.create(PATH_OUT_ANALYSIS) }
```

```{r load-data, include=FALSE}
# LOAD DATA
fltd <- readRDS(PATH_RDS_FILE)

# LOAD CLUSTERING RESULTS
clusters <- read.table(PATH_CLUSTER_TABLE, header = TRUE, sep = ",", row.names = "cellIDs")
clustOfInterest <- grep(paste0(DefaultAssay(fltd), "_snn_res.", clust_res), names(clusters), value = TRUE)
fltd@meta.data <- merge(fltd@meta.data, clusters[clustOfInterest], by = 'row.names')
fltd@meta.data <- tibble::column_to_rownames(fltd@meta.data, "Row.names")
```

`r if(atlas){cat("# Cells annotation through label transfer from the atlas")}`

```{r mapping-atlas, eval=atlas, child="60_cellTypeAnnotationTransfer.Rmd"}
```

<!-- ## Impact of the additional controls on the annotation transfer -->

<!-- ```{r heatmap-afterDF} -->
<!-- Idents(fltd) <- fltd@meta.data$celltype_DF -->
<!-- transition <- table(fltd@meta.data$celltype_DF, fltd@meta.data$celltype_singleData) -->
<!-- dt <- as.data.frame(transition, row.names = names(fltd@meta.data$celltype_singleData)) -->
<!-- dt[dt == 0] <- NA -->
<!-- names(dt) <- c("beforeDF", "afterDF", "counts") -->


<!-- ggplot(dt, aes(beforeDF, afterDF, fill = counts)) + # nolint -->
<!--     geom_tile() + -->
<!--     geom_text(aes(label = counts), color = "black", size = 2.5, fontface = "bold") + -->
<!--     #scale_color_manual(values = c(NA, "black"), na.value = NA) + -->
<!--     coord_fixed() + -->
<!--     scale_fill_gradient(low = "white", high = "pink", na.value="white") + -->
<!--     theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) + -->
<!--     guides(fill = guide_colourbar(barwidth = 0.5, -->
<!--                                   barheight = 10, -->
<!--                                   title = "Counts")) + -->
<!--     ggtitle(paste0(DATASET, " : impact of doublets removal on the label transfer")) + -->
<!--     CenterTitle() + -->
<!--     labs(x = "Cell type annotation before additional controls", y = paste0("Cell type annotation after additional controls")) -->
<!-- ``` -->



```{r write-ensemblID-mapping-section, results='asis', eval=(gn_col == 1), echo=(gn_col==1)}
cat("

# Map ensembl IDs to gene names

Presentation of Biomart, selection of the reference genome.
")

ensembl.ids <- rownames(fltd)

# # explore the biomart database
# listEnsembl()
# ensembl <- useEnsembl(biomart = "genes")
# datasets <- listDatasets(ensembl)
# listEnsemblArchives() # useful to retrieve an old annotation

# selection of the reference genome for annotation
mm10 <- useEnsembl(biomart = 'genes', 
                   dataset = 'mmusculus_gene_ensembl',
                   version = 108) # ==> GRCm38.p3 reference genome

geneNames108 <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"), #, "external_transcript_name"),
                     filters = "ensembl_gene_id",
                     values = ensembl.ids,
                     mart = mm10)

# Check for manually inserted genes
manInsert <- grep("ENS", ensembl.ids, value = TRUE, invert = TRUE)
geneNames108 <- rbind(geneNames108, as.data.frame(cbind(ensembl_gene_id = manInsert, external_gene_name = manInsert)))

write.table(geneNames108, file = file.path(PATH_OUT_ANALYSIS, "correspondanceENSid_GeneName_V108.csv"), col.names = TRUE, row.names = FALSE, sep = ",")

duplicatesGN <- geneNames108[which(duplicated(geneNames108$external_gene_name) | duplicated(geneNames108$external_gene_name, fromLast = TRUE)),]
```

```{r reduce-gene-list, eval=(gn_col == 1), results='asis'}
cat(paste0("
The ", DATASET," dataset has **", length(ensembl.ids), "** different feature genes
comprising ", length(manInsert), " manually inserted genes:  
", toString(manInsert), "

***

Number of retrieved ensembl gene IDs: 
**", length(unique(geneNames108$ensembl_gene_id)), "**  
*(the manually inserted genes are already added to the retrieved data)*

List of the **", length(setdiff(ensembl.ids, geneNames108$ensembl_gene_id)), "**
not found ensembl gene IDs:  
", toString(setdiff(ensembl.ids, geneNames108$ensembl_gene_id)), "

***

Number of unique retrieved gene names: **", length(unique(geneNames108$external_gene_name)), "**

"))
```


```{r table-duplicateGN, echo=FALSE, eval=gn_col == 1 && dim(duplicatesGN)[1] !=0, results='asis'}
cat("

Some gene names fit multiple ensembl gene IDs. Below you can see the table of
the redundant gene names and the corresponding ensembl gene IDs.
")
# datatable(topNmarkers, rownames = FALSE, filter = "top", options = list(pageLength = 5, scrollX=T))
if(gn_col == 1){
duplicatesGN %>%
    rename(gene_names = external_gene_name) %>%
    arrange(gene_names) %>%
    knitr::kable(row.names = FALSE, align = "ll") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, position = "center") %>%
  column_spec(1, width = "12em") %>%
  column_spec(2, width = "12em")
}

cat("
<br>

We removed all ensembl gene IDs mapped to multiple gene names. We reduce the
feature list to the remaining ensembl gene IDs.
")
```

```{r rename-genes-seurat, eval=(gn_col == 1), echo=(gn_col==1)}
# Remove ensembl gene IDs with multiple match to gene names
if (dim(duplicatesGN)[1] !=0) {
    geneNamesUniq <- geneNames108[!geneNames108$ensembl_gene_id %in% duplicatesGN$ensembl_gene_id, ]
} else {
    geneNamesUniq <- geneNames108
}

SOsubset <- subset(fltd, features = geneNamesUniq$ensembl_gene_id)

# Function to rename genes in seurat object
renameFeatures <- function(SO = NULL, assay = "RNA", newNames = NULL) {

	# Focus on a specific assay
	obj <- SO@assays[[assay]]

	# Replacement labels must have the same length with the data
	if (length(newNames$external_gene_name) == nrow(obj)) {

		# Replace the names in the counts and data matrices
		if (length(obj@counts)) obj@counts@Dimnames[[1]]    <- newNames$external_gene_name
        if (length(obj@data)) obj@data@Dimnames[[1]]        <- newNames$external_gene_name
	
		# Replace the names in the scales matrix
		if (length(obj@scale.data)) dimnames(obj@scale.data)[[1]] <- newNames$external_gene_name

		# Replace the names in the variable features matrices
		if (length(obj@var.features) > 0) {

			# rename variable features with gene names
			ensemblFeat <- data.frame(hvg = obj@var.features)
    	    geneFeat <- ensemblFeat %>%
				left_join(newNames, by = join_by(hvg == ensembl_gene_id))
    	    obj@var.features <- geneFeat$external_gene_name

			# change rownames of the variable features meta data table
			# WORKS ONLY WHEN DUPLICATES ARE REMOVED
			tmp.meta.features <- obj@meta.features %>%
				tibble::rownames_to_column("ensembl_gene_id") %>%
    	        right_join(newNames, ., by = "ensembl_gene_id") %>%
    	        tibble::column_to_rownames("external_gene_name") %>%
    	        select(-(ensembl_gene_id))
			obj@meta.features <- tmp.meta.features
		}

	} else {
	   cat("Error: Unequal gene sets: nrow(RNA) != nrow(newnames)")
	}

	# Return the object
	SO@assays[[assay]] <- obj

	return(SO)
}


fltd <- renameFeatures(SOsubset, DefaultAssay(SOsubset), geneNamesUniq)
if (DefaultAssay(fltd) == "integrated") fltd <- renameFeatures(fltd, "RNA", geneNamesUniq)

fltd
```


# Perform differential expression analysis

```{r run-dea}
clustOfInterest <- grep(paste0(DefaultAssay(fltd), "_snn_res.", clust_res), names(fltd@meta.data), value = TRUE)
Idents(fltd) <- clustOfInterest

DefaultAssay(fltd) <- 'RNA'

markers <- FindAllMarkers(fltd, only.pos = TRUE)
#markers.sign <- markers[markers$p_val_adj < 0.001 & markers$avg_logFC > log(1.5), ]
markers.sign <- markers[order(-markers$avg_log2FC), ]
markers.sign$cluster <- factor(markers.sign$cluster, levels = as.character(sort(as.numeric(levels(markers.sign$cluster)))))
markers.sign <- markers.sign[order(markers.sign$cluster), ]
markers.sign <- markers.sign[ , c(6, 7, 2:4, 1, 5)]
write.table(x = markers.sign, file = file.path(PATH_OUT_ANALYSIS,
                                               paste0("markers_min.pct_logFC_default_", 
                                                      DATASET, "_scenario.", scenario, "_res.", clust_res, ".csv")), sep = ",", row.names = F, col.names = T)

topNmarkers <- extract_top_features(markers.sign, topn = top_markers)


write.table(x = topNmarkers, file = file.path(PATH_OUT_ANALYSIS,
                                              paste0("top_", top_markers, "_markers_", DATASET, "_scenario.", scenario, "_res.", clust_res, ".csv")), sep = ",", row.names = F, col.names = T)

datatable(topNmarkers, rownames = FALSE, filter = "top", options = list(pageLength = 5, scrollX=T))
```


# Additional Plots

```{r load-goi-list, eval=!is.null(genes_of_interest), echo=!is.null(genes_of_interest)}
# Load list of genes of interest
goi <- try(read.table(genes_of_interest), silent = TRUE)

if (inherits(goi, "try-error")) {
    goi <- read.table(genes_of_interest, sep=",", colClasses=c("character", "NULL"))
}

goiList <- goi$V1

genesToRemove <- goiList[!(goiList %in% rownames(fltd))]
if (length(genesToRemove) >= 1){
    cat(paste0("One or more genes of interest are not in the dataset (maybe because of the case):\n\t",
               paste(genesToRemove, collapse = ", "), "\nThe listed genes will not be plotted"))
}

goiList <- goiList[goiList %in% rownames(fltd)] # remove the unknown genes

```

## Cluster identity {.tabset .tabset-pills .tabset-fade}

```{r get-topNmarkers}
# Parameters
ntop <- 5

topNmarkers <- extract_top_features(markers.sign, topn = ntop)
topNgenes <- unique(topNmarkers$gene)
Idents(fltd) <- factor(fltd@meta.data[,clustOfInterest], levels = sort(as.numeric(levels(as.factor(fltd@meta.data[,clustOfInterest])))))
```

### DimPlot of the cluster identity
```{r cluster-DimPlot-markers, out.width="50%"}
p <- DimPlot(fltd, reduction = "umap", pt.size = 0.8,
             label = FALSE) +
  ggtitle(paste0("Clusters at resolution ", clust_res, "\n", DATASET)) +
  CenterTitle() +
  NoAxes()
LabelClusters(p, id = "ident",  fontface = "bold", size = 5)
```

### DotPlot of the markers on cluster identity

```{r cluster-DotPlot-markers}
create_dot_plot(fltd, topNgenes, title=paste0("Top ", ntop, " markers of each cluster in clusters"))
```


### DotPlot of the genes of interest on cluster identity

```{r cluster-DotPlot-goi}
if(!is.null(genes_of_interest)){

create_dot_plot(fltd, sort(goiList), title="Genes of interest in clusters")
} else {
  cat("No genes of interest provided, nothing to see here.")
}
```

### FeaturePlots of the markers {.tabset .tabset-pills .tabset-fade}

```{r cluster-FeaturePlots-markers, results='asis', out.width='20%', fig.align='default'}
# code with the courtesy of Lionel Spinelli and Romain Fenouil
topMarkers = by( markers.sign, markers.sign[["cluster"]], function(x)
{
  # Filter markers based on adjusted PValue
  x = x[ x[["p_val_adj"]] < 0.005, , drop = FALSE]
  # Sort by decreasing logFC
  x = x[ order(abs(x[["avg_log2FC"]]), decreasing = TRUE), , drop = FALSE ]
  # Return top ones
  return( if(is.null(  top_markers )) head( x, n =  10 ) else head( x, n =  min(top_markers, 10) ))
})

clustersColor <- setNames(hue_pal()(length(names(topMarkers))), names(topMarkers))

invisible( lapply( names( topMarkers), function(clusterName)
{
  cat("#### Cl. <span style='border-radius: 3px border: 3px solid ", clustersColor[clusterName], " padding:0px 2px'>", clusterName, "</span>\n")
  
  # Highlight cells of current cluster on a dimreduc plot
  highlightClusterPlot(clusterName, seuratObject = fltd, reduction = ifelse( exists("useReduction"), useReduction, "umap"))
  
  # Plots expression on projected cells
  invisible( lapply( topMarkers[[clusterName]][["gene"]], function(featureName)
  {
    print( FeaturePlot( fltd, features = featureName, reduction = ifelse( exists("useReduction"), useReduction, "umap"), order = TRUE) +
            theme( axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    legend.position = "none"))
  }))
  
  cat(" \n \n") # Required for '.tabset'
}))
```

### FeaturePlots of the genes of interest with cluster labels

```{r cluster-FeaturePlot-goi, out.width='20%', fig.align='default'}
if(!is.null(genes_of_interest)){
# FeaturePlot the genes of interest
FeaturePlot( fltd,
    features = goiList, 
    reduction = ifelse( exists("useReduction"), useReduction, "umap"),
    min.cutoff = "q01", max.cutoff = "q99",
    order = TRUE, label = TRUE,
    combine = FALSE)

} else {
  cat("No genes of interest provided, nothing to see here.")
}
```

### ViolinPlots of the markers {.tabset .tabset-pills .tabset-fade}

```{r clusters-VlnPlots-markers, results='asis', out.width='20%', fig.align='default'}
# with courtesy of Lionel Spinelli and Romain Fenouil
invisible( lapply( names( topMarkers), function(clusterName)
{
  cat("#### Cl. <span style='border-radius: 3px border: 3px solid ", clustersColor[clusterName], " padding:0px 2px'>", clusterName, "</span>\n")
  
  # Remind cluster name in an empty figure to keep consistent alignment of panels between tabs
  plot( c( 0, 1), c( 0, 1), ann = F, bty = 'n', type = 'n', xaxt = 'n', yaxt = 'n')
  text( x = 0.5, y = 0.5, paste( "Cluster", clusterName), cex = 2, col = clustersColor[clusterName])
  
  # Violinplot for expression value of marker genes by cluster (+ number of 'zero' and 'not zero' cells)
  invisible( lapply( topMarkers[[clusterName]][["gene"]], violinFeatureByCluster, seuratObject = fltd, clustersColor = clustersColor))
  
  cat(" \n \n") # Required for '.tabset'
}))
```

```{r set-celltype-identity, eval=atlas}
cat("## Celltype identity {.tabset .tabset-pills .tabset-fade}")
Idents(fltd) <- factor(fltd@meta.data[, colname], levels = sort(levels(as.factor(fltd@meta.data[, colname]))))
```

```{r celltype-DimPlot, eval=atlas, results='asis', out.width="50%"}
cat("### DimPlot of the celltype identity {.tabset .tabset-pills .tabset-fade}")
## DimPlot the celltype
DimPlot(fltd,
		reduction = "umap",
        pt.size = 1,
        repel = TRUE,
        label = TRUE,
        label.size = 6,
        cols = colors.celltype[levels(Idents(fltd))]) +
		ggtitle(paste0(DATASET, " : cell identities")) +
		defaultTheme() +
		NoLegend()

cellsData <- data.frame(fltd@reductions[["umap"]]@cell.embeddings, fltd@meta.data[, colname])
colnames(cellsData) <- c(colnames(fltd@reductions[["umap"]]@cell.embeddings), "celltype")

invisible(sapply(levels(as.factor(fltd@meta.data[, colname])), function(act)
{
    cat("#### ", act, "\n")
        
    print(ggplot(cellsData[c('UMAP_1', 'UMAP_2')], # Omit the column used for facetting to get all points repeated in all facets
         aes( x = UMAP_1,
              y = UMAP_2)) +
    geom_point( alpha = .4,
                size  = 1,
                color = "grey") +
    geom_point( data = cellsData[cellsData$celltype == act,], # Now provide data including column for facetting
                aes(color = celltype),
                alpha = .5,
                size  = 3.5) +
    blankTheme() +
    scale_fill_manual(values = colors.celltype[fltd@meta.data[, colname]], aesthetics = "color") +
    NoLegend() +
    ggtitle(paste0(act))
    )

    cat(" \n \n") # Required for '.tabset'
}))
```

```{r celltype-DotPlot-markers, eval=atlas}
cat("### DotPlot of the markers on celltype identity")
## DotPlot the n-top markers of each celltype
create_dot_plot(fltd, topNgenes, title=paste0("Top ", ntop, " markers of each cluster in celltypes"))
```

```{r celltype-DotPlot-goi, eval=atlas}
cat("### DotPlot of the genes of interest on celltype identity")
if(!is.null(genes_of_interest)){
## DotPlot the genes of interest
create_dot_plot(fltd, sort(goiList), title="Genes of interest in celltypes")
} else {
  cat("These are not the droids you are looking for.")
}
```


<!-- ## Personal exploration -->

<!-- List of genes obtained from the CHD-RF-KB (A knowledgebase for non-syndromic -->
<!-- congenital heart disease associated risk factors). -->

<!-- ```{r} -->
<!-- # Load list of genes of interest -->
<!-- goi <- read.table(file.path(PATH_REQUIREMENTS, "geneList_CHD-RF-KB.txt")) -->
<!-- goiList <- goi$V1 -->
<!-- goiList <- str_to_title(goiList) -->

<!-- genesToRemove <- goiList[!(goiList %in% rownames(fltd))] -->
<!-- if (length(genesToRemove) >= 1){ -->
<!--   cat(paste0("One or more genes of interest are not in the dataset (maybe because of the case):\n\t", -->
<!--              paste(genesToRemove, collapse = ", "), "\nThe listed genes will not be plotted")) -->
<!-- } -->

<!-- goiList <- goiList[goiList %in% rownames(fltd)] # remove the unknown genes -->

<!-- Idents(fltd) <- factor(fltd@meta.data[,clustOfInterest], levels = sort(as.numeric(levels(as.factor(fltd@meta.data[,clustOfInterest]))))) -->
<!-- dotPlotsGoi(fltd, goiList, title="Genes of CHD-RF-KB database") -->
<!-- ``` -->


# SessionInfo

```{r sessionInfo}
sessionInfo()
```