---
author: "Céline Chevalier and Anaı̈s Baudot"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
# title: |
#   | Doublets removal
#   | Step {{STEP_ID}}
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type = "text/javascript">
  $(document).ready(function() {
    $("body").prepend("<div class = \"zoomDiv\"><img src = \"\" class = \"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use "img.zoom" and out.extra = "class = \"zoom\"" in chunk to specify manually which chunk images can be zoomed
    $("img:not(.zoomImg)").filter(":not(a *)").click(function() {
      $(".zoomImg").attr("src", $(this).attr("src"));
      $(".zoomDiv").show();
    })
    // onClick function for hiding div
    $("img.zoomImg").click(function() {
      $(".zoomDiv").hide();
    })
  })
</script>


```{r setup04-1, include = FALSE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,

  # text output
  echo = TRUE,
  results = "hold",
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,

  # code decoration
  tidy.opts = list(width.cutoff = 90),
  comment = "",
  attr.output = ".numberLines",

  # plots
  fig.path = paste0(PATH_OUT_FIG, "/07_combine_ata_"),
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c("png", "pdf"),
  out.width = "50%",
  fig.width = 12,
  fig.height = 12,
  fig.align = "center"   # should be tuned to default in multiple plots chunk
)
```

```{r load_lib}
library(Seurat)
library(ggplot2)

source("./plotting_functions.R")
source("./data_management.R")
```

```{r load-colors, include = FALSE}
# Create color vectors
colors.table <- read.csv(file.path(PATH_ATLAS, "colorsSheet.csv"), header=T, comment.char="", as.is=T)
colors.celltype <- setNames(colors.table$celltype_color[!is.na(colors.table$celltype)], colors.table$celltype[!is.na(colors.table$celltype)])
colors.stage <- setNames(colors.table$development_stage_color[!is.na(colors.table$development_stage)], colors.table$development_stage[!is.na(colors.table$development_stage)])
```

```{r variableDef1, include=FALSE}
# Split comma separated input files into a list
PATH_OUT_ANALYSIS <- file.path(PATH_ROOT, "07_annotDEAviz")
if (!dir.exists(PATH_OUT_ANALYSIS)) { dir.create(PATH_OUT_ANALYSIS) }
```

```{r load-data, include=FALSE}
# Load data
data_load <- function(DATASET) {
    # PATHS
    PATH_RDS_FILE <- paste0(PATH_RDS_OBJECTS, "/05_prepocessed_", DATASET, "_scenario.", scenario, ".rds")
    # PATH_CLUSTER_TABLE <- paste0(PATH_ROOT, "/06_process/clustering_", DATASET, "_scenario.", scenario, "_meth.", clust_meth, "_res.", clust_res, ".csv")
    
    data <- readRDS(PATH_RDS_FILE)
}

SO.list <- lapply(input_datasets, data_load)
names(SO.list) <- input_datasets
```

```{r integrate-sep-data, eval=combine_meth!="merge", echo=combine_meth!="merge"}
set.seed(general_seed)

if (combine_meth == "blkH"){
    SO <- merge(x = SO.list[[1]], y = SO.list[2:length(SO.list)], merge.data = TRUE)
    
} else if (combine_meth == "blkS"){
    features <- SelectIntegrationFeatures(object.list = SO.list)
    anchors <- FindIntegrationAnchors(object.list = SO.list, anchor.features = features)
    SO <- IntegrateData(anchorset = anchors)
    
} else if (combine_meth == "seqS"){
    SO <- SO.list[[1]]
    names(SO.list) <- 
    for (a_day in names(SO.list)[2:length(SO.list)]){
        integ.list <- list(SO, SO.list[[a_day]])
        features <- SelectIntegrationFeatures(integ.list)
        anchors <- FindIntegrationAnchors(object.list = integ.list,
                                          anchor.features = features)
        SO <- IntegrateData(anchorset = anchors)
    } 
}

rm(SO.list)
invisible(gc())
```


```{r merge-sep-data, eval=combine_meth=="merge", echo=combine_meth=="merge"}
SO <- merge(x = SO.list[[1]], y = SO.list[2:length(SO.list)], merge.data = T)
SO

rm(SO.list)
invisible(gc())
```


SHOW THE TABLE OF SO$ORIG.IDENT TO CHECK THE NUMBER OF CELLS FROM EACH DATASET

# Preprocessing workflow

The preprocessing workflow consists in the steps of

- data normalization,

- data centering,

- highly variable genes identification,

- principal components analysis (PCA),

- non-linear dimension reduction for visualization, and

- clustering.

```{r preprocess-1, eval=hvg_meth != "mvp", echo=hvg_meth != "mvp", message=TRUE}
SO <- ScaleData(SO, features=rownames(SO), do.scale=do_scale, verbose=FALSE)
SO <- FindVariableFeatures(SO, nfeatures=hvg_num, selection.method = hvg_meth, verbose=FALSE)
var.feats.dataset <- VariableFeatures(SO)
#saveRDS(var.feats.dataset, file=file.path(rdsObjects, paste0("var.feats.dataset_", dataset, "_", combine_meth, ".rds")))

SO <- RunPCA(SO,
             npcs = pca_npcs,
             nfeatures.print = pca_print,
             seed.use = general_seed,
             verbose=TRUE)

if (combine_meth == "blkH"){
   	SO <- RunHarmony(SO, "orig.ident", plot_convergence = FALSE)
}

SO <- RunUMAP(SO,
              reduction = ifelse( combine_meth == "blkH", "harmony", "pca"),
              dims = 1:top_pcs,
              seed.use = general_seed,
              verbose = FALSE)
SO <- FindNeighbors(SO,
                    reduction = ifelse( combine_meth == "blkH", "harmony", "pca"),
                    dims = 1:top_pcs,
                    verbose=FALSE)
SO <- FindClusters(SO,
                   resolution = clust_res,
                   algorithm = clust_meth,
                   random.seed = general_seed,
                   verbose=FALSE)
```


```{r hvg-mvp2, eval=hvg.meth == "mvp", echo=hvg.meth == "mvp", message=TRUE}
SO <- ScaleData(SO, features=rownames(SO), do.scale=do_scale.scale, verbose=FALSE)

SO <- FindVariableFeatures(SO, selection.method = hvg.hvg_meth, verbose=FALSE)
head(SO@assays[[DefaultAssay(SO)]]@meta.features) %>%
  knitr::kable(caption = "Example of HVG selection output") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

t(as.data.frame(table(SO@assays[[DefaultAssay(SO)]]@meta.features$mvp.variable))) %>%
  knitr::kable(caption = "Number of features selected as highly variable genes (HVG)",
               row.names = FALSE,
               escape = F,
               align = "cc") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

var.feats.dataset <- VariableFeatures(SO)
#saveRDS(var.feats.dataset, file=file.path(rdsObjects, paste0("var.feats.dataset_", dataset, "_", combine_meth, ".rds")))

SO <- RunPCA(SO, npcs = pca.pca_npcs, nfeatures.print = pca.pca_print, seed.use = general_seed, verbose=TRUE)

if (combine_meth == "blkH"){
    SO <- RunHarmony(SO, "orig.ident", plot_convergence = FALSE)
}

SO <- RunUMAP(SO,
              reduction = ifelse( combine_meth == "blkH", "harmony", "pca"),
              dims = 1:top_pcs,
              seed.use = general_seed,
              verbose = FALSE)
SO <- FindNeighbors(SO,
                    reduction = ifelse( combine_meth == "blkH", "harmony", "pca"),
                    dims = 1:top_pcs,
                    verbose=FALSE)
SO <- FindClusters(SO,
                   resolution = clust_res,
                   algorithm = clust_meth,
                   random.seed = general_seed,
                   verbose=FALSE)
```

```{r save2}
saveRDS(SO, file=file.path(PATH_RDS_OBJECTS, paste0("06_combine_", DATASET, "_scenario.", scenario, "_method.", combine_meth, ".rds")))
invisible(gc())
```


```{r callAtlas}
fltd <- SO
rm(SO)
invisible(gc())
```

```{r mapping-atlas, child="60_cellTypeAnnotationTransfer.Rmd"}
```
