---
author: "Céline Chevalier and Anaı̈s Baudot"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        theme:
            bootswatch: yeti
        toc: yes
        toc_float:
          collapsed: false
          smooth_scroll: true
        number_sections: yes
        df_print: kable
        code_folding: hide
    pdf_document:
        number_sections: yes
        toc: yes
        toc_depth: 3
        keep_tex: no
# title: |
#   | Doublets removal
#   | Step {{STEP_ID}}
---

<!-- Javascript for zooming on figures (adapted from: https://stackoverflow.com/questions/40401680) -->

<!-- Jquery import conflicts with DT::datatable so needs to be commented here -->
<!-- <script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.1.1/jquery.min.js"></script> -->

<style>
.zoomDiv {
  display: none;  
  position: fixed;
  top: 50%;
  left: 50%;
  z-index: 50;
  transform: translate(-50%, -50%);
  background-color: #FFFFFF;
  box-shadow: 0px 0px 50px #888888;
  width: fit-content;
  max-width: 90%;
  max-height: 90%;
  overflow: auto;
}

.zoomImg {
  width: 150%;
}
</style>

<script type = "text/javascript">
  $(document).ready(function() {
    $("body").prepend("<div class = \"zoomDiv\"><img src = \"\" class = \"zoomImg\"></div>");
    // onClick for all img except the zoomed one and link ones (filter)
    // use "img.zoom" and out.extra = "class = \"zoom\"" in chunk to specify manually which chunk images can be zoomed
    $("img:not(.zoomImg)").filter(":not(a *)").click(function() {
      $(".zoomImg").attr("src", $(this).attr("src"));
      $(".zoomDiv").show();
    })
    // onClick function for hiding div
    $("img.zoomImg").click(function() {
      $(".zoomDiv").hide();
    })
  })
</script>


```{r setup, include = FALSE}
options(knitr.purl.inline = TRUE)
knitr::opts_chunk$set(
  # code evaluation
  eval = TRUE,

  # text output
  echo = TRUE,
  results = "hold",
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  strip.white = TRUE,

  # code decoration
  tidy.opts = list(width.cutoff = 90),
  comment = "",
  attr.output = ".numberLines",

  # plots
  fig.path = paste0(PATH_OUT_FIG, "/06_annotDEAviz_scenario_", scenario, "_"),
  fig.show = "asis",         # tuned to "hold" in multiple plots chunk
  dev = c("png", "pdf"),
  out.width = "50%",
  fig.width = 12,
  fig.height = 12,
  fig.align = "center"   # should be tuned to default in multiple plots chunk
)
```

```{r load-libraries, include = FALSE}
library(Seurat)
library(ggplot2)
#library(biomaRt)
library(tidyr)
library(Matrix)
library(dplyr)
library(kableExtra)
library(stringr)
library(DT)
library(scales)
library(RColorBrewer)
library(grDevices)
library(DElegate)

# source("./extract_top_features.R")
# source("./highlightClusterPlot.R")
# source("./additionalPlots.R")
source("./plotting_functions.R")
source("./data_management.R")
```